The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-03-12 src/ms.tex
add scipy as dependency
First draft of the Finance Section (#31)
* First draft of funding section

* Fixed minor issue with a reference

* Newline added

* Rview comments addresssed

* added reference

As part of the process of developing \astropypkg into a long-term sustainable
product, and to improve transparency and accountability, the Project agreed to
write down and formalize our governance structure (partly supported by explicit
funding for this purpose; see Section~\ref{sec:project-funding}). At the 2019
Astropy Coordination Meeting, input was gathered from participants on what governance
structures existed in the associated Open Source Software communities, and what
would fit well with the needs of \astropy. This led into a ``retreat'' planned
for March 2020, but due to the COVID-19 pandemic, this became a series of
virtual meetings of the ``Astropy Governance Working Group.'' This group
drafted the APE 0 document \citep{ape0}, which was then eventually ratified and
implemented by the ``Astropy Governance Implementation Working Group'' in Fall
2021. While the process emphasized flexibility and the ability to adapt to
changing circumstances, it is expected that this is the framework Astropy's
governance will operate in for at least the medium-term future.

The APE0 \citep{ape0} document lays out the principles of this governance
structure, so we refer the reader to that document for a more thorough
description, and here we only highlight some key elements. While many of these
principles were already de facto true or have been discussed organically (and
have been discussed in earlier papers in this series), the APE0-based governance
aims to provide a single place where the community can agree as a starting
point. With this in mind, it highlights the developer and user community of the
Astropy Project as the ultimate sources of authority, as well as the core
principle of ``do-ocracy'' that those who do work for the Project (be it coding,
training, or other harder-to-quantify contributions) gain more influence on the
outputs of the Project by virtue of their effort. However, APE0 adds the
concept of ``voting members'': a self-governed part of that community who are
entrusted to elect the Coordination Committee (CoCo). While the CoCo has existed
from the inception of the project, APE0 establishes a formal voting process for
CoCo members, and explicitly outlines the rights and responsibilities of the
CoCo. This role is mainly to facilitate consensus and act as the decision maker
when other mechanisms have failed, and includes powers that either
require central authority or secrets (e.g., passwords) However, APE0 also
charges the Committee to devolve responsibilities and seek community input on
these items as often as possible.

The first Coordination Committee election under these rules took place in Fall
2021, electing a mix of prior and new coordination committee members,
and was contested in the sense of more candidates than available slots. This
suggests the process is already working to serve the long-term interests of the
committee to both spread the coordination effort, and to ensure it is not
dominated by the same people for as long as the Project continues. While other,
more fine-grained governance improvements are planned for the future, it is
clear the foundation is now in place.

\subsection{Core package contributor base} \label{sec:project-contributors}

@@

% TODO: move this subsection to "updates to core package" section above?

\begin{figure}[t!]
    \begin{centering}
      \includegraphics[width=0.8\textwidth]{Ncommitters_vs_frac_commits.pdf}
        \caption{
            The cumulative fraction of Git commits to the \astropypkg core
            package as a function of number of contributors (committers) broken
            into three equal-length time periods between the start of \astropy
            and the feature freeze date of v5.0.
            The time eras early, mid, and recent correspond to the approximate
            year ranges 2011--2015, 2015--2018, 2018--2021 (defined more
            precisely in Section~\ref{sec:project-contributors}).
            In all eras, the majority of commits to the \astropypkg core package
            have come from a small subset of the total number of contributors.
            For example, as annotated in the figure, 90\% of the Git commits
            have come from 18, 34, and 44 committers in the early, mid, and
            recent eras, respectively.
        }
        \label{fig:contributor-summary:Ncommitters}
    \end{centering}
\end{figure}

As described in the introduction above, the \astropy Project is broader than the
core \astropypkg package as it also represents a number of core infrastructure
packages, affiliated packages, educational initiatives, and user communities.
In total, well over 1500 people have contributed in some way to the Project,
whether, for example, through contributing code or documentation, opening issues
on \github, providing infrastructure support, writing tutorials, or helping to
coordinate workshops that relate to Astropy.
The vast majority of these contributors and participants have interacted as
volunteers choosing to engage with the open-source and open-development software
community.
However, while the total number of contributors is large, the number of
contributions and bulk of the maintenance of the \astropy Project components has
primarily been sustained by a small number of heavily-engaged or
explicitly-employed contributors \citep[see, e.g., ][]{astropy:2018}, as is the
case in many large, open-source software communities \citep[e.g.,][]{numpy}.

As an example of this within the \astropy Project, here we focus on
contributions made to the \astropypkg core package.
Figure~\ref{fig:contributor-summary:Ncommitters} shows the number of
contributors who have contributed a given fraction of the total number of Git
commits, separated into three separate time spans (``eras'') in the history of
the Project.\footnote{Note, however, that measuring these statistics in terms of
Git commits is not a perfect way of encapsulating ``effort,'' but we use this
metric for its simplicity and clear definition.}
The annotations show that, within the early (July 2011--January 2015), mid
(January 2015--June 2018), and recent (June 2018--November 2021) eras, 90\% of
the commits to the \astropypkg core package have come from 18, 34, and 44
contributors.
While the trend is encouraging --- a larger pool of contributors are
contributing more, fractionally --- this emphasizes that, still, the vast
majority of the work in \astropypkg is done my a small subset ($<20\%$) of the
115, 216, and 234 contributors, respectively, within each era.

\begin{figure}[t!]
    \begin{centering}
      \includegraphics[width=0.75\textwidth]{top_contributors_career.pdf}
        \caption{
            todo
        }
        \label{fig:contributor-summary:top}
    \end{centering}
\end{figure}

Since its inception, the most active contributors to the \astropypkg package
have spanned a range of career stages and career paths.
For example, Figure~\ref{fig:contributor-summary:top} shows a breakdown of the
career stage/path at the midpoint of the three time eras defined in the previous
paragraph for all contributors with more than 50 commits to the \astropypkg
package within each era.
Over the $\sim$6 years between the ``early'' and ``recent'' eras, many
contributors moved between these categories (i.e., have graduated as students to
become research staff / faculty or software engineers), but also a number of
early contributors moved on to other positions and stopped contributing to the
core \astropypkg package.
Given this, it is encouraging that the numbers in each career stage/path have
stayed relatively constant over the history of \astropy (with some showing
growth), as new, active contributors have generally joined and stepped in to
fill top contributor roles within the project.
On the other hand, the total number of top contributors has not grown
substantially, reflecting one of our principal challenges: Sustaining a pipeline
from users to contributors to maintainers of the package.
One other thing to note in this context is that in all eras, 50--60\% of the
software engineers who contributed to \astropypkg were (or are) employed by
STScI, whose leadership have consistently provided resources (financial and
personnel) towards sustaining \astropypkg.

\begin{figure}[t!]
    \begin{centering}
      \includegraphics[width=0.75\textwidth]{commits_committers_per_month.pdf}
        \caption{
            todo
        }
        \label{fig:contributor-summary:per-month}
    \end{centering}
\end{figure}

A final metric we consider ... rate of contributions and contributor participation ...

@@

% Much of this text was adapted from Robel Geda's report from NSBP and
% SACNAS. Please include him on the author list.
\textbf{\astropy representation at national diversity conferences:}
Several members of the \astropy community attended virtual conferences of the
National Society of Black Physicists (NSBP) and SACNAS (focusing on Hispanic and
Native American scientists across all STEM fields). \astropy representatives
noted that one underlying topic came up multiple times. A major barrier for
persons who come from underrepresented communities is the lack of resources and
expertise that are necessary to train students. Writing PEP 8 compliant software
and understanding the Git and \github workflow is not part of the standard
Physics and Astrophysics curriculum. Additionally, the larger astronomical
community is still in the process of transitioning towards \python-based tools,
making it difficult for students not currently under advisement by some one with
extensive \python expertise to get involved with the open source \python
community. Projects such as Learn Astropy (see Section~\ref{sec:learn}) could
have a profound impact by empowering underrepresented groups because it provides
a free, searchable, and accessible introduction to \python tools for
astrophysical research. Representatives also noted that offering training and
teaching materials to PIs of Research Experience for Undergraduate (REU)
programs throughout the United States could be helpful for encouraging advisors
to teach students how to use Astropy, create their own libraries, and use
version control on open source platforms like \github. Such materials could also
be offered as a workshop at national diversity conferences themselves, as SACNAS
solicits special session proposals each year.

% \textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
%   initiative:} \todo{No author assigned}{Describe more about this.}

\subsection{Project funding} \label{sec:project-funding}

% \secauthor{Aarya Patil}

@@

With the goal of establishing a sustainable financial model for the Astropy
Project, the Coordination Committee, under the new governance charter (see
Section~\ref{sec:project-governance}), established a standing Finance Committee.
This committee oversees the planning and allocation of finances on behalf
of the Project. The Astropy Project accepts funds from institutions as well as
individuals through NumFOCUS.\footnote{NumFOCUS is a 501(c)(3) nonprofit that
supports and promotes world-class, innovative, open source scientific computing.}
Previously, no direct financial support was available for the project development,
and NumFOCUS covered most of the incurred operational costs. However, transitioning
to long-term financial stability meant having an influx of funds in the form of
institutional support, and hence the Astropy Project applied for and successfully
acquired four grants over the last several years. These grants are administered by NumFOCUS on behalf of the
Project, providing a ``neutral'' space not tied to any particular astronomy
institution. The two major grants are briefly summarized below.

\begin{description}
    \item[Moore] The Gordon and Betty Moore Foundation awarded the Astropy
    project a $\sim$\$900k (US) grant in 2019 for the proposal ``Sustaining and
    Growing the Astropy Project.'' Over a three-year term, this grant supported
    sustenance of the Project by providing funds for (1) the development and
    maintenance of the \astropypkg and its infrastructure, provided to existing
    project members and targeted hires, (2) the transition of long-term users to
    contributors and maintainers through mentorship efforts, (3) the
    formalization of a governance structure (see
    Section~\ref{sec:project-governance}), and (4) the improvement of equity,
    diversity, and inclusion efforts in the project (see
    Section~\ref{sec:project-ide}). Some of this money supported travel for
    meetings, conferences, and workshops. This generous grant helped reduce
    reliance on volunteer-driven maintenance of the openly-developed package and
    paved the way to acquire funding from federal agencies, representing a major
    milestone for the success of the project.

    \item[NASA] A successful proposal to the NASA ROSES-2020
    call\footnote{http://solicitation.nasaprs.com/ROSES2020}, section E.7
    (Support for Open Source Tools, Frameworks, and Libraries), granted the
    Astropy Project $\sim$\$600k (US) to support its work. Funded work for a
    term of three years includes (1) project-wide infrastructure maintenance and
    improvements, (2) targeted work on areas of the \astropy core and
    coordinated packages, (3) enhancements to the Learn
    Astropy ecosystem and educational materials (see Section~\ref{sec:learn}),
    and (4) support for the \astropy affiliated packages. These goals
    support the sustenance of the project by encouraging further
    development of the package ecosystem and community engagement.
\end{description}

In addition to the two grants described above, the project benefited from
a Gemini Observatory contract under its Science User Support Department,
which awarded funds for development work that supports both the Astropy and
DRAGONS projects. The project also received financial support from the Dunlap
Seed Funding program at the University of Toronto to develop educational
resources for scientific software packaging within the Learn Astropy
framework. All of these successful proposals show that the Astropy Project
is vital for the astronomical infrastructure and community at large, and
strengthen its promising and sustainable future.

@@

% \secauthor{Nicholas Earl}

The Astropy project supports the broader ecosystem by providing pre-configured
infrastructure packages that the community can use to support and maintain their
own software package infrastructure. These include tools and extensions that
enable easily generating documentation and setting up automated testing, as well
as providing package scaffolding for new projects.

Sphinx is a common and useful tool for generating documentation for Python
packages. The Astropy project maintains a default Sphinx configuration along
with Astropy-specific extensions that can be easily added to community
projects via the
\href{https://github.com/astropy/sphinx-astropy}{\texttt{sphinx-astropy}} meta
package. This tool provides a pre-configured Sphinx setup compatible with
Astropy projects, which includes several extensions useful for generating API
documentation
(\href{https://github.com/astropy/sphinx-automodapi}{\texttt{sphinx-automodapi}}),
allowing for Numpy docstring parsing
(\href{https://github.com/numpy/numpydoc}{\texttt{numpydoc}}), embedded image
handling
(\href{https://github.com/sphinx-gallery/sphinx-gallery}{\texttt{sphinx-gallery}};
\href{https://github.com/python-pillow}{\texttt{pillow}}), advanced
documentation testing support
(\href{https://github.com/astropy/pytest-doctestplus}{\texttt{pytest-doctestplus}}),
and providing a custom documentation theme ideal for analysis packages
(\href{https://github.com/astropy/sphinx-astropy}{\texttt{astropy-sphinx-theme}}).

Community package testing infrastructure is supported through the
\href{https://github.com/astropy/pytest-astropy}{\texttt{pytest-astropy}}
meta-package, providing a unified testing framework with useful extensions
compatible with both Astropy- and non-Astropy-affiliated community packages.
This meta-package pulls in several
\href{https://github.com/pytest-dev/pytest}{\texttt{pytest}} plugins to help
with custom test headers
(\href{https://github.com/astropy/pytest-astropy-header}{\texttt{pytest-astropy-header}}),
accessing remotely-hosted data files in tests
(\href{https://github.com/astropy/pytest-remotedata}{\texttt{pytest-remotedata}}),
interoperability with documentation
(\href{https://github.com/astropy/pytest-doctestplus}{\texttt{pytest-doctestplus}}),
dangling file handle checking
(\href{https://github.com/astropy/pytest-openfiles}{\texttt{pytest-openfiles}}),
data array comparison support in tests
(\href{https://github.com/astropy/pytest-arraydiff}{\texttt{pytest-arraydiff}}),
sub-package command-line testing support
(\href{https://github.com/astropy/pytest-filter-subpackage}{\texttt{pytest-filter-subpackage}}),
improved mock object testing
(\href{https://github.com/pytest-dev/pytest-mock}{\texttt{pytest-mock}}), test
coverage reports and measurements
(\href{https://github.com/pytest-dev/pytest-cov}{\texttt{pytest-cov}}), and
configuring packages for property-based testing
(\href{https://github.com/HypothesisWorks/hypothesis}{\texttt{hypothesis}}).

The Astropy Package Template helps facilitate the setup and creation of new
Python packages leveraging the Astropy ecosystem. This tool utilizes the
Cookiecutter project to walk users through the process of creating new
packages complete with documentation and testing support. Additionally, the
package template generation process includes the ability to setup
interoperability with \github, allowing for easy repository access from
documentation, as well as an example \github Actions workflow to demonstrate
the use of \github's continuous integration tooling.

\subsection{Astropy affiliated packages}

% \secauthor{Matt Craig, Brett Morris}

% Highlight a few new affiliated packages and major updates to existing ones.
% Include a big table of all affiliated packages and references (as in v2.0
% paper).

As defined in \paperii, Astropy affiliated packages are astronomy-related
\python packages that provide functionality that builds upon and extends the
\astropypkg core package that have requested to be included as part of the
Astropy Project community. These packages support the goals and vision of
Astropy of improving code re-use, interoperability, and embracing good coding
practices such as testing and thorough documentation.
All Astropy-coordinated and other affiliated packages are listed with detailed
information on the \href{{https://www.astropy.org/affiliated}}{\astropy
website}; A table summarizing key aspects of the affiliated packages is included
in the Appendix below (Table~\ref{tab:affiliated-registry}).

Since \paperii, there has been an expansion of affiliated
packages for gravitational astrophysics, including:
\texttt{PyCBC} for exploring gravitational wave signals, \texttt{lenstronomy} for
modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing
gravitational wave probability maps, and \texttt{EinsteinPy} for general
relativity and gravitational astronomy. There have also been several packages
added to the ecosystem related to HEALPix: \texttt{astropy-healpix}
for a BSD-licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order
Coverage maps. \texttt{astroalign} has been introduced for astrometric registration,
and \texttt{python-cpl} has been added for ESO pipelines and VLT data products.
For ground-based astronomy, \texttt{baseband} has added IO capabilities for
VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the \astropy ecosystem.
Other new affiliated packages include \texttt{agnpy} for AGN jets,
\texttt{statmorph} for fitting galactic morphological diagnostics,
\texttt{dust\_extinction} for modeling interstellar dust extinction,
\texttt{feets} for extracting features from time series data, and
\texttt{corral} for managing data intensive parallel pipelines.
\texttt{saba} gives an interface to the Sherpa \citep{sherpa} fitting routines,
\texttt{BayesicFitting} provides an interface for generic Bayesian inference,
and \texttt{sbpy} enables calculations for asteroid and
cometary astrophysics. Finally, \texttt{synphot} provides an interface for
synthetic photometry.
% \citep{Mommert2019}

Several of the coordinated and affiliated packages described in
\cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
\citep{astroquery} has added access to roughly a dozen new missions and data
services, including the James Webb Space Telescope (JWST) archive, and the
project has switched to a continuous release model: Every time a change is
committed to the main development branch it is published on the Python Package
Index (PyPI) and available for installation. (However, formal releases are still
done a few times per year.) \texttt{photutils} \cite {photutils} released its
first stable version, indicating that the API will change less frequently, and
there have been several significant performance improvements. \texttt{ccdproc}
\cite{ccdproc} also released a new major version, bringing better performance to
some image combination operations. The \texttt{regions} package
\cite{pyregions}, for manipulating ds9-style region definitions \cite{ds9},
added new ways to manipulate regions and introduced new region types.
\texttt{reproject}'s \cite{reproject} major new feature is a function to align
and co-add images to create a mosaic; better support for parallelization was
also added. \texttt{specutils}, the package that defines containers for 1D and
2D spectra \cite{specutils}, also had its first stable release and the addition
of classes to read JWST data. \texttt{stingray} has also had a major performance
overhaul. The package \texttt{gammapy} \cite{gammapy} also has major performance
improvements and has unified its API in preparation for its first stable
release.

\subsection{Connections with data archives}
\label{sec:astroquery}

% \secauthor{Adam Ginsburg}

\texttt{astroquery} \citep{Ginsburg2019} is the Astropy-coordinated package for
interacting with online archives of astronomical and related data. It contains
over 50 modules for querying astronomical databases, large and small. In
particular, since \paperii, significant contributions to \texttt{astroquery}
have come from several of the major archives, including the European Space
Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space
Telescope Science Institute (STScI), the Infrared Science Archive at the NASA
Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian
Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter
Array archive (ALMA). These contributions represent a formal acknowledgement of
the utility of a centralized tool suite for archive interaction from \python.
The widespread usage of \texttt{astroquery} is apparent from the range of
keywords represented in the journal articles that cite the \texttt{astroquery}
paper \citep{Ginsburg2019}; everything from asteroids to galaxies is
represented. Additionally \texttt{astroquery} functionality has been built into
several special-purpose \python packages, such as \texttt{LightKurve}
\citep{LightKurve} and SORA \citep{SORA}]. More than 4,000 repositories on
\github make use of \texttt{astroquery} in some way, and \texttt{astroquery}
features in a large number of tutorials on various facets of astronomical
analysis.

Many of the existing and newly-contributed tools rely on Virtual Observatory
(VO) tools. These use the underlying package \texttt{pyvo}, which has also
recently become an Astropy-coordinated package. While many or all of the
functions provided in \texttt{astroquery} can be achieved through direct use of
VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more
closely resemble the web interfaces that are more familiar to most users.

@@

In this section, we highlight a few efforts in observatory- or mission-driven
development that have contributed to Astropy, and vice versa.

\subsubsection{James Webb Space Telescope}
% \secauthor{Larry Bradley}

The James Webb Space Telescope (JWST) is a 6.5-meter space-based
infrared telescope that will provide unprecedented resolution and
sensitivity from 0.6--28 microns. JWST will enable a broad range
of scientific investigations from exoplanets and their atmospheres
to the formation of galaxies in the very early universe. Its four
key scientific goals are to study the first light from stars and
galaxies, the assembly and evolution of galaxies, the birth of stars and
protoplanetary systems, and planetary systems and the origins of life.

@@

% Gemini
% \secunfilled

>>>>

As part of the process of developing Astropy into a long-term sustainable
product, and to improve transparency and accountability, the Project agreed to
write down and formalize our governance structure (partly supported by explicit
funding for this purpose - see \S \ref{sec:project-funding}). At the 2019
Astropy Coordination Meeting, input was gathered from participants on what governance
structures existed in the associated Open Source Software communities, and what
would fit well with the needs of Astropy. This led into a ``retreat'' planned
for March 2020, but due to the COVID-19 pandemic, this became a series of
virtual meetings of the ``Astropy Governance Working Group''.  This group
drafted the APE 0 document \citep{ape0}, which was then eventually ratified and
implemented by the ``Astropy Governance Implementation Working Group'' in Fall
2021. While the process emphasized flexibility and the ability to adapt to
changing circumstances, it is expected that this is the framework Astropy's
governance will operate in for at least the medium-term future.

The APE0 \citep{ape0} document lays out the principles of this governance
structure, so we refer the reader to that document for a more thorough
description.  However here we highlight some key elements. While many of these
principles were already de facto true or have been discussed organically (and
have been discussed in earlier papers in this series), the APE0-based governance
aimed to provide a single place where the community can agree as a starting
point. With this in mind, it highlights the developer and user community of the
Astropy Project as the ultimate sources of authority, as well as the core
principle of ``do-ocracy'' that those who do work for the Project (be it coding,
training, or other less concrete contributions) gain more influence on the
outputs of the Project by virtue of their effort.  APE0 adds, however, the
concept of ``voting members'' - a self-governed part of that community who are
entrusted to elect the Coordination Committee. While this committee has existed
from the inception of the project, APE0 establishes a formal voting process for
this committee, and explicitly outlines the rights and responsibilities of the
Coordination Committee. This role is mainly to facilitate consensus and act as
the decision maker when other mechanisms have failed.  However, it also includes
powers that either require central authority or secrets (e.g., passwords), but
APE0 also charges the Committee to devolve responsibilities and seek community
input on these items as often as possible.

The first Coordination Committee election under these rules took place in Fall
2021, electing a mix of prior existing and new coordination committee members,
and was contested in the sense of more candidates than available slots. This
suggests the process is already working to serve the long-term interests of the
committee to both spread the coordination effort, and to ensure it is not
dominated by the same people for as long as the Project continues. While other,
more fine-grained governance improvements are planned for the future, it is
clear the foundation is now in place.

\subsection{Contributor base} \label{sec:project-contributors}

@@

Overview and statistics of contributors. Highlight changes since v2.0.

\begin{figure}
    \begin{centering}
      % \includegraphics{figures/contributor-summary.pdf}
        \caption{Placeholder figure!}
        \label{fig:contributor-summary}
    \end{centering}
\end{figure}

@@

% Much of this text was adapted from Robel Geda's report from NSBP and
% SACNAS. Please include him on the author list.
\textbf{\astropy representation at national diversity conferences:}
Several members of the \astropy community attended virtual conferences
of the National Society of Black Physicists (NSBP) and SACNAS
(focusing on Hispanic and Native American scientists across all STEM
fields). \astropy representatives noted that one underlying topic came
up multiple times. A major barrier for persons who come from
underrepresented communities is the lack of resources and expertise
that are necessary to train students. Writing PEP 8 compliant software
and understanding the Git and Github workflow is not part of the
standard Physics and Astrophysics curriculum. Additionally, the larger
Astronomnical community is still in the process of transitioning
towards Python based tools, making it difficult for students not
currently under advisement by some one with extensive Python expertise
to get involved with the open source Python community. Projects such
as Learn Astropy could have a profound impact by empowering
underrepresented groups because it provides a free, searchable, and
accessible introduction to Python tools for astrophysical
research. Representatives also noted that offering training and
teaching materials to PIs of Research Experience for Undergraduate
(REU) programs throughout the United States could be helpful for
encouraging advisors to teach students how to use Astropy, create
their own libraries, and use version control on open source platforms
like Github. Such materials could also be offered as a workshop at
national diversity conferences themselves, as SACNAS solicits special
session proposals each year.

\textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
  initiative:} \todo{No author assigned}{Describe more about this.}

\subsection{Funding} \label{sec:project-funding}

\secauthor{Aarya Patil}

@@

With the goal of establishing a sustainable financial model for the Astropy
Project, the Coordination Committee under the new governance charter (see \S
\ref{sec:project-governance}) established a standing Finance committee.
This committee oversee the planning and allocation of finances on behalf
of the project. The Astropy Project accepts funds from institutions as well as
individuals through NUMFOCUS\footnote{NUMFOCUS is a 501(c)(3) nonprofit that
supports and promotes world-class, innovative, open source scientific computing.}.
Previously, no direct financial support was available for the project development,
and NUMFOCUS covered most of the incurred operational costs. However, transitioning
to long-term financial stability meant having an influx of funds in the form of
institutional support, and hence the Astropy Project applied for and successfully
acquired four grants. These grants are administered by NumFOCUS on behalf of the
project, providing a ``neutral'' space not tied to any particular astronomy
institution. The two major ones are briefly summarized below.

\begin{itemize}
    \item Moore: The Gordon and Betty Moore Foundation awarded the Astropy
    project a $\sim$\$900k (US) grant in 2019 for the proposal ``Sustaining and
    Growing the Astropy Project". Over a three-year term, this grant supported
    sustenance of the project by providing funds for (1) the development and
    maintenance of the \astropypkg and its infrastructure to existing project
    members and targeted hires, (2) the transition of long-term users to
    contributors and maintainers through mentorship efforts, (3) the formalization
    of a governance structure (see \S \ref{sec:project-governance}), and (4) the
    improvement of equity, diversity, and inclusion efforts in the project (see \S
    \ref{sec:project-ide}). Some of the money supported travel for meetings,
    conferences, and workshops. This generous grant helped reduce reliance on
    volunteer-driven maintenance of the openly-developed package and paved
    the way to acquire funding from federal agencies: a big milestone for the
    success of the project.

    \item NASA: A successful proposal to the NASA ROSES-2020
    call\footnote{http://solicitation.nasaprs.com/ROSES2020}, section E.7
    (Support for Open Source Tools, Frameworks, and Libraries), granted the
    Astropy Project $\sim$\$600k (US) to support its work. Funded work for a
    term of three years includes (1) project-wide infrastructure maintenance and
    improvements, (2) targeted work on areas of the \astropy core and
    coordinated packages that need support, (3) enhancements to the Learn
    Astropy ecosystem and educational materials (see \S \ref{sec:learn}),
    and (4) support for the \astropy affiliated packages. These goals
    support the sustenance of the project by encouraging further
    development of the package ecosystem and community engagement.
\end{itemize}

In addition to the two grants described above, the project benefited from
a Gemini Observatory contract under its Science User Support Department,
which awarded funds for development work that supports both the Astropy and
DRAGONS projects. The project also received financial support from the Dunlap
Seed Funding program at the University of Toronto to develop educational
resources for scientific software packaging within the Learn Astropy
framework. All these successful proposals show that the Astropy project
is vital for the astronomical infrastructure and community at large, and
strengthen its promising and sustainable future.

@@

\secauthor{Nicholas Earl}

The Astropy project supports the broader ecosystem by providing pre-configured
infrastructure packages that the community can use to support and maintain
their own software package infrastructure. These include tools to easily
generate documentation and setup automated testing, as well as provide
package scaffolding for new projects.

Sphinx is a common and useful tool for generating documentation for Python
packages. The Astropy project maintains a default Sphinx configuration along
with Astropy-specific extensions which can be easily added to community
projects via the
\href{https://github.com/astropy/sphinx-astropy}{\texttt{sphinx-astropy}} meta
package. This tool provides a pre-configured Sphinx setup compatible with
Astropy projects, which includes several extensions useful for generating API
documentation
(\href{https://github.com/astropy/sphinx-automodapi}{\texttt{sphinx-automodapi}}),
allowing for Numpy docstring parsing
(\href{https://github.com/numpy/numpydoc}{\texttt{numpydoc}}), embedded image
handling
(\href{https://github.com/sphinx-gallery/sphinx-gallery}{\texttt{sphinx-gallery}};
\href{https://github.com/python-pillow}{\texttt{pillow}}), advanced
documentation testing support
(\href{https://github.com/astropy/pytest-doctestplus}{\texttt{pytest-doctestplus}}),
and providing a custom documentation theme ideal for analysis packages
(\href{https://github.com/astropy/sphinx-astropy}{\texttt{astropy-sphinx-theme}}).

Community package testing infrastructure is supported through the
\href{https://github.com/astropy/pytest-astropy}{\texttt{pytest-astropy}}
meta-package, providing a unified testing framework with useful extensions
compatible with both Astropy- and non-Astropy-affiliated community packages.
This meta-package pulls in several
\href{https://github.com/pytest-dev/pytest}{\texttt{pytest}} plugins to help
with custom test headers
(\href{https://github.com/astropy/pytest-astropy-header}{\texttt{pytest-astropy-header}}),
accessing remotely-hosted data files in tests
(\href{https://github.com/astropy/pytest-remotedata}{\texttt{pytest-remotedata}}),
interoperability with documentation
(\href{https://github.com/astropy/pytest-doctestplus}{\texttt{pytest-doctestplus}}),
dangling file handle checking
(\href{https://github.com/astropy/pytest-openfiles}{\texttt{pytest-openfiles}}),
data array comparison support in tests
(\href{https://github.com/astropy/pytest-arraydiff}{\texttt{pytest-arraydiff}}),
sub-package command-line testing support
(\href{https://github.com/astropy/pytest-filter-subpackage}{\texttt{pytest-filter-subpackage}}),
improved mock object testing
(\href{https://github.com/pytest-dev/pytest-mock}{\texttt{pytest-mock}}), test
coverage reports and measurements
(\href{https://github.com/pytest-dev/pytest-cov}{\texttt{pytest-cov}}), and
configuring package for property-based testing
(\href{https://github.com/HypothesisWorks/hypothesis}{\texttt{hypothesis}}).

The Astropy Package Template helps facilitate the setup and creation of new
Python packages leveraging the Astropy ecosystem. This tool utilizes the
Cookiecutter project to walk users through the process of creating new
packages complete with documentation and testing support. Additionally, the
package template generation process includes the ability to setup
interoperability with GitHub, allowing for easy repository access from
documentation, as well as an example GitHub Actions workflow to demonstrate
the use of GitHub's continuous integration tooling.

\subsection{Affiliated packages}

\secauthor{Matt Craig, Brett Morris}

Highlight a few new affiliated packages and major updates to existing ones.
Include a big table of all affiliated packages and references (as in v2.0
paper).

All coordinated and affiliated packages are listed on the astropy
website\footnote{\url{https://www.astropy.org/affiliated}}. Since 
\cite{astropy:2018}, there has been an expansion of affiliated
packages for gravitational astrophysics including:
\texttt{PyCBC} for exploring gravitational wave signals, \texttt{lenstronomy} for
modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing
gravitational wave probability maps, and \texttt{EinsteinPy} for general
relativity and gravitational astronomy. There have been several packages
added to the ecosystem related to HEALPix: \texttt{astropy-healpix}
for a BSD-licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order
Coverage maps. \texttt{astroalign} has been introduced for astrometric registration,
and \texttt{python-cpl} has been added for ESO pipelines and VLT data products. 
For ground-based astronomy, \texttt{baseband} has added IO capabilities for 
VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the astropy ecosystem. 
We have \texttt{agnpy} for AGN jets, and \texttt{statmorph} for fitting galactic
morphological diagnostics. \texttt{saba} gives an interface to sherpa's fitting 
routines. For modeling interstellar dust extinction we have 
\texttt{dust\_extinction}, and for extracting features from time series we
have \texttt{feets}. We also have \texttt{corral} for managing data intensive parallel 
pipelines. \texttt{BayesicFitting} provides an interface for generic Bayesian 
inference, and \texttt{sbpy} \cite{Mommert2019} does calculations for asteroid and cometary
astrophysics. Finally, \texttt{synphot} provides an interface for synthetic photometry.

Several of the coordinated and affiliated packages described in \cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
\cite{astroquery} has added
access to roughly a dozen new missions and data services, including JWST, and
the project has switched to a continuous release model. Every time a change
is committed to the main development branch it is published on the Python
Package Index (PyPI, ref) and available for installation. Formal releases are
still done a couple of times per year. \texttt{photutils} \cite
{photutils} released its first stable version, indicating that the API will
change less frequently, and there have been several significant performance
improvements.  \texttt{ccdproc} \cite{ccdproc} also released a new major
version, bringing better performance to some image combination operations.
The \texttt{regions} package \cite{pyregions}, for manipulating ds9-style
region definitions \cite{ds9}, added new ways to manipulate regions and
introduced new region types. \texttt{reproject}’s \cite{reproject} major new
feature is a function to align and co-add images to create a mosaic; better
support for parallelization was also added. \texttt{specutils}, the package
that defines containers for 1D and 2D spectra \cite{specutils}, also had its
first stable release and the addition of classes to read JWST data.
\texttt{stingray} has also had a major performance overhaul. The package
\texttt{gammapy} \cite{gammapy} also has major performance improvements,
has unified its API in preparation for its first stable release.

\subsection{Connections with data archives}

\label{sec:astroquery}
% MHvK: feel free to move above label to where it makes most sense.

\secauthor{Adam Ginsburg}

Astroquery \cite{Ginsburg2019} is the astropy-coordinated package for interacting with online archives of astronomical and related data. It contains over 50 modules for querying astronomical databases, large and small. In particular, since \paperii, significant contributions to \texttt{astroquery} have come from several of the major archives, including the European Space Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space Telescope Science Institute (STScI), the Infrared Science Archive at the NASA Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter Array archive (ALMA). These contributions represent a formal acknowledgement of the utility of a centralized tool suite for archive interaction.

The widespread usage of \texttt{astroquery} is apparent from the range of keywords represented in the journal articles which cite \cite{Ginsburg2019}; everything from astroids to galaxies is represented. Additionally Astroquery functionality has been built into several special purpose Python packages, such as \texttt{LightKurve} \cite{LightKurve} and SORA \cite{SORA}]. More than 4,000 repositories on GitHub make use of \texttt{astroquery} in some way, an \texttt{astroquery} features in a large number of tutorials on various facets of astronomical analysis.

Many of the existing and newly-contributed tools rely on Virtual Observatory (VO) tools.
These use the underlying package \texttt{pyvo}, which has also recently become an astropy-coordinated package. While many or all of the functions provided in \texttt{astroquery} can be achieved through direct use of VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more closely resemble the web interfaces more familiar to users.
% note to coauthors: we need a section on pyvo

@@

Brief summary of efforts in observatory- or mission-driven development that have
contributed to Astropy, and vice versa.

\subsubsection{James Webb Space Telescope}
\secauthor{Larry Bradley}

The James Webb Space Telescope (JWST) is a 6.5-meter space-based
infrared telescope that will provide unprecedented resolution and
sensitivity from 0.6 -- 28 microns. JWST will enable a broad range
of scientific investigations from exoplanets and their atmospheres
to the formation of galaxies in the very early universe. Its four
key scientific goals are to study the first light from stars and
galaxies, the assembly and evolution of galaxies, the birth of stars and
protoplanetary systems, and planetary systems and the origins of life.

@@

Gemini
\secunfilled

Cherenkov Telescope Array
\secauthor{Axel Donath, Maximilian Nöthe}
